---
title: "Analysis of GPT-3 Humor Appreciation"
author: "Drew Walker, Sean Trott, Seana Coulson"
date: "July 19, 2023"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "pdf")
```


```{r include=FALSE}
library(tidyverse)
library(lmtest)
library(forcats)
library(broom)
library(lme4)
library(ggridges)
library(lmerTest)
```

# Load data

## Load classification task

```{r}
## setwd("/Users/seantrott/Dropbox/UCSD/Research/NLMs/humor_llms/src/analysis")
df_class = read_csv("../../data/processed/llm_classification.csv")
nrow(df_class)

head(df_class)
```


## Load appreciation task

```{r}
## setwd("/Users/seantrott/Dropbox/UCSD/Research/NLMs/humor_llms/src/analysis")
df_app = read_csv("../../data/processed/llm_appreciation.csv")
nrow(df_app)

head(df_app)

```


## Load surprisals

```{r}
## setwd("/Users/seantrott/Dropbox/UCSD/Research/NLMs/humor_llms/src/analysis")
df_surprisals = read_csv("../../data/processed/results_gpt.csv")
df_surprisals_td02 = df_surprisals %>%
  filter(model == "text-davinci-002")
```


# Classification task

## Merge with surprisals

```{r}
df_class_merged = df_class %>%
  inner_join(df_surprisals)

nrow(df_class_merged)
```


## `Log Odds ~ Is Joke`

Overall, there is a very clear effect: GPT-3 assigns higher probability to the "yes" answer (as in, *yes*, this is a joke) for jokes than non-jokes.

```{r}
mod_full = lm(data = df_class_merged, lp_ratio ~ is_joke)
summary(mod_full)
```

Though interestingly, GPT-3 seems biased towards a *no* response (i.e., log-odds is generally negative):

```{r}
df_class_merged %>%
  group_by(is_joke) %>%
  summarise(m = mean(lp_ratio),
            sd = sd(lp_ratio))
```


We can also visualize this:

```{r}
df_class_merged %>%
  ggplot(aes(x = lp_ratio,
             y = is_joke,
             fill = is_joke)) +
  geom_density_ridges2(aes(height = ..density..), 
                       color=gray(0.25), 
                       alpha = 0.5, 
                       scale=0.85, 
                       size=.9, 
                       stat="density") +
  labs(x = "GPT-3 Log-odds (yes vs. no)",
       y = "Joke vs. Non-Joke",
       fill = "Joke vs. Non-Joke") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme_minimal() +
  # scale_fill_viridis_d() +
  theme(
    legend.position = "bottom"
  ) + 
  theme(axis.title = element_text(size=rel(2)),
        axis.text = element_text(size = rel(2)),
        legend.text = element_text(size = rel(2)),
        legend.title = element_text(size = rel(2)),
        strip.text.x = element_text(size = rel(2)))


df_class_merged %>%
  ggplot(aes(x = is_joke,
             y = lp_ratio)) +
  stat_summary (fun = function(x){mean(x)},
                fun.min = function(x){mean(x) - 2*sd(x)/sqrt(length(x))},
                fun.max = function(x){mean(x) + 2*sd(x)/sqrt(length(x))},
                geom= 'pointrange', 
                position=position_dodge(width=0.95),
                size = .5, alpha = .5) +
  labs(x = "Joke vs. Non-joke",
       y = "Log Odds (Yes vs. No)") +
  theme_minimal() 

```


# Appreciation task

## Merge with surprisals

```{r}
df_app_merged = df_app %>%
  inner_join(df_surprisals)

nrow(df_app_merged)
```


```{r}
mean(df_app_merged$best_response)
df_app_merged %>%
  ggplot(aes(x = best_response)) +
  geom_histogram(bins = 5, alpha = .5) +
  theme_minimal() +
  labs(x = "Response") 
```



## `Funniness ~ Is Joke + Surprisal`

```{r}
mod_full = lm(data = df_app_merged, best_response ~ is_joke + surprisal)
summary(mod_full)
```

Both factors explain variance.

```{r}
df_app_merged %>%
  group_by(is_joke) %>%
  summarise(m = mean(best_response),
            sd = sd(best_response))

df_app_merged %>%
  ggplot(aes(x = is_joke,
             y = best_response)) +
  stat_summary (fun = function(x){mean(x)},
                fun.min = function(x){mean(x) - 2*sd(x)/sqrt(length(x))},
                fun.max = function(x){mean(x) + 2*sd(x)/sqrt(length(x))},
                geom= 'pointrange', 
                position=position_dodge(width=0.95),
                size = .5, alpha = .5) +
  labs(x = "Joke vs. Non-joke",
       y = "Funniness Rating (GPT-3)") +
  theme_minimal() 

df_app_merged %>%
  ggplot(aes(x = best_response,
             y = surprisal,
             color = is_joke)) +
  stat_summary (fun = function(x){mean(x)},
                fun.min = function(x){mean(x) - 2*sd(x)/sqrt(length(x))},
                fun.max = function(x){mean(x) + 2*sd(x)/sqrt(length(x))},
                geom= 'pointrange', 
                # position=position_dodge(width=0.95),
                size = .5, alpha = .5) +
  labs(x = "Funniness Rating (GPT-3)",
       y = "Surprisal",
       color = "Joke vs. Non-joke") +
  theme_minimal() 


```


# Exploratory

## Classification

```{r}
mod_full = lm(data = df_class_merged, lp_ratio ~ is_joke * surprisal)
summary(mod_full)


df_class_merged %>%
  ggplot(aes(x = surprisal, y = lp_ratio, color = is_joke)) +
  geom_point(alpha = .3) +
  geom_smooth(method = "lm") +
  theme_minimal() +
  labs(x = "Surprisal",
       y = "Log Odds (Yes vs. No)",
       title = "Joke Classification") +
  facet_wrap(~is_joke)
```


## `Entropy ~ Is Joke + Surprisal`

```{r}
mod_full = lm(data = df_app_merged, entropy ~ is_joke * surprisal)
summary(mod_full)


df_app_merged %>%
  ggplot(aes(x = surprisal, y = entropy, color = is_joke)) +
  geom_point(alpha = .3) +
  geom_smooth(method = "lm") +
  theme_minimal() +
  labs(x = "Surprisal",
       y = "Entropy over Funniness Responses",
       title = "Funniness Responses") +
  facet_wrap(~is_joke)

```


## Select items

For joke detection (LLMs):

```{r}
df_class_merged %>%
  filter(is_joke == FALSE) %>%
  filter(lp_ratio > 0) %>%
  select(sentence, condition, lp_ratio, surprisal)
```



